---
title: "AssociationRules"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(arules)
library(ggplot2)
library(xtable)
```

## Reading the transaction file
Our transaction file is given in the "basket" format, which means that each line represents a transaction, with the transaction ID as first column and the next columns being objects contains in this transaction (in this context: buying of each customer).
```{r}
ttr <- read.transactions("Transactions.txt", format="basket", header=FALSE, sep=" ", cols=1)
```

## First exploration
```{r}
summary(ttr)
head(as(ttr, "list"))
```

### What is the most frequent item ? (without filtering)
```{r}
headFreq = head(sort(itemFrequency(ttr), decreasing = TRUE), n=12)
headFreqDf = data.frame(keyName=names(headFreq), value=headFreq, row.names=NULL)
print(xtable(headFreqDf, type="latex"), file="tables\\mfitset_beforeapriori.tex")
barplot(headFreq, cex.names=0.7)
```
The most frequent item is the 1104010101, present in more than half of purchases (0.51) followed by 1301120101 and 1104020101. The maxlen parameters has not been applied yet but will be for the association rules analysis.


## Apriori application
```{r}
ap <- apriori(ttr, parameter=list(support=0.001, confidence=0.1, maxlen=5))
summary(ap)
```

### Most frequent itemsets (after filtering)
```{r}
fsets <- unique(generatingItemsets(ap))
fsets.df <- as(fsets, "data.frame")
sor.fsets <- fsets.df[order(-fsets.df$support),]
sor.fsets[1:12,]
print(xtable(sor.fsets[1:12,], type="latex"), file="tables\\mfitset_afterapriori.tex")
```
The 10 most frequent \footnote{Equivalent to sort itemset by their support} itemsets are single-item sets, which means that they are more likely to appear than any other itemset (single or multiple). It is also interesting to notice that the filtering we applied on the data (using support, confidence and maxlen parameters) had no influence on the top 10 items (being itemsets after apriori).

However, if we dig further, we can see that the 12th most frequent itemset is containing 2 items (1301120101,1301120200 with a support of 0.26) which is actually composed of the 2nd and 5th most frequent (single-item) itemsets.

## Sorted rules
### Rules sorted by confidence

### Rules sorted by lift
Let's know comapre rules in the lift criteria which quantifies how much better the rule is than a random prediction. A lift higher than 1 means that the rule is indeed reliable to make a prediction when a lift lower than 1 means that a random prediction is better than the rule, which makes it non-relevent.
```{r}
lifted <- sort(ap, by="lift")
inspect(lifted[1:3])
print(xtable(inspect(lifted[1:3]), type="latex"), file="tables\\top_rules_lift.tex")
```
The top 3 rules sorted by lift can be seen in the table below:  

TABLE

We can see that those three rules have a very high lift (around 150) which means that we are able to predict the right-hand side element (for example 1302140700 for the first rule) using this rule.





